---
title: "Homework-SDS"
author: "Andrea De Vincenzo, Domenico Azzarito and Michele Pezza"
date: "`r Sys.Date()`"
output:
  html_document:
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

In this part, we work on deriving and understanding the Bayes
classification rule, denoted as $\eta^*(x)$, in a simple binary
classification scenario. The goal of this rule is to minimize
classification errors by assigning each observation to the class with
the higher posterior probability.

Here, we assume $Y \in \{0, 1\}$ is a binary outcome, and
$X \in \mathbb{R}$ is a continuous feature. The conditional
distributions of $X$ given $Y$ are defined as uniform distributions, and
both classes have equal prior probabilities. Using this setup, we derive
the regression function $r(x)$, which gives the conditional probability
$\mathbb{P}(Y=1 \mid X=x)$, and use it to define the decision rule.

In this section, we break down the steps to calculate $\eta^*(x)$ by
comparing the conditional densities $f_0(x)$ and $f_1(x)$ for the two
classes. This leads to a clear definition of the decision boundary,
showing how the optimal rule can be applied to classify observations
based on their feature values.

## 1. Bayes classification rule $\eta^{*}(x)$

$(Y,X)$ are random variables with $Y \in \{0,1\}$ and
$X \in \mathbb{R}$. Suppose that

$$
(X \mid Y = 0) \sim \text{Unif}(-3, 1) \quad \text{and} \quad (X \mid Y = 1) \sim \text{Unif}(-1, 3)
$$

Further suppose that
$\mathbb{P}(Y = 0) = \mathbb{P}(Y = 1) = \frac{1}{2}$.

------------------------------------------------------------------------

The regression function is defined as follows: $$
r(x) = \mathbb{E}(Y \mid X = x) = \mathbb{P}(Y=1 \mid X = x) = \dfrac{\pi_1f_1(x)}{\pi_1f_1(x) + (1-\pi_1)f_0(x)}
$$

where
$\pi_1 = \mathbb{P}(Y = 1), f_1(x) = f(x \mid Y = 1) \text{ and }  f_0(x) = f(x \mid Y = 0)$.

------------------------------------------------------------------------

The Bayes classification rule $\eta^{*}(x)$ is defined as:

$$
\eta^*(x) = 
\begin{cases} 
1 & \text{if } \mathbb{P}(Y = 1 | X = x) > \mathbb{P}(Y = 0 | X = x) \\
0 & \text{otherwise}
\end{cases} = \begin{cases}
1 & \text{if } \pi_1f_1(x) > (1-\pi_1)f_0(x) \\
0 & \text{otherwise}
\end{cases}
$$

Since we have $\pi_1 = \pi_0 = \frac{1}{2}$: $$
\eta^*(x) = 
\begin{cases} 
1 & \text{if } f_1(x) > f_0(x) \\
0 & \text{otherwise}
\end{cases}
$$

------------------------------------------------------------------------

In our setup: $$
f_1(x) = 
\begin{cases} 
\frac{1}{4} & \text{if } -1 \leq x \leq 3 \\
0 & \text{otherwise}
\end{cases} 
\quad \text{and} \quad f_0(x) = 
\begin{cases} 
\frac{1}{4} & \text{if } -3 \leq x \leq 1 \\
0 & \text{otherwise}
\end{cases} 
$$

------------------------------------------------------------------------

-   $x < -3$: $$
    f_1(x) = 0, \; f_0(x) = 0 \Rightarrow f_1(x) \not> f_0(x) \Rightarrow \eta^{*}(x) = 0 
    $$
-   $-3 \leq x < -1$: $$
    f_1(x) = 0, \; f_0(x) = \frac{1}{4} \Rightarrow f_1(x) \not> f_0(x) \Rightarrow \eta^{*}(x) = 0 
    $$
-   $-1 \leq x \leq 1$: $$
    f_1(x) = \frac{1}{4}, \; f_0(x) = \frac{1}{4} \Rightarrow f_1(x) \not> f_0(x) \Rightarrow \eta^{*}(x) = 0 
    $$
-   $1 < x \leq 3$: $$
    f_1(x) = \frac{1}{4}, \; f_0(x) = 0 \Rightarrow f_1(x) > f_0(x) \Rightarrow \eta^{*}(x) = 1 
    $$
-   $x > 3$: $$
    f_1(x) = 0, \; f_0(x) = 0 \Rightarrow f_1(x) \not> f_0(x) \Rightarrow \eta^{*}(x) = 0
    $$

Hence, here the Bayes classification rule $\eta^{*}(x)$ is:

$$
\eta^*(x) = 
\begin{cases} 
1 & \text{if} \quad 1 < x \leq 3 \\
0 & \text{otherwise}
\end{cases}
$$

## 2.1

Now, we generate a dataset of size $n = 1000$ from the joint data model
$p(y, x) = p(x | y) \cdot p(y)$ described earlier. The data will include
samples drawn from the specified uniform distributions for $X$
conditioned on $Y$. We will then visualize the generated dataset.

```{r point 1.2, echo=TRUE}
source('Functions.R')
set.seed(54) # We begin with setting the seed for reproducibility
n <- 1000 # Sample size 
# Define the distribution

# We simulate one dataset
data <- simulate(n)  # defined in 'Functions.R'

# Plot the data
plot(data$x, data$y, 
     col = ifelse(data$y == 1, adjustcolor("darkred", alpha.f = 0.1), 
                                  adjustcolor("darkblue", alpha.f = 0.1)), 
     pch = 16, cex = 0.7, xlab = "X", ylab = "Y", 
     main = "Generated dataset and Regression function", 
     sub = expression(eta(x) == 1 ~ "if and only if" ~ r(x) > 0.5))

# Plot the Regression Function r(x)
x_seq <- seq(-3, 3, length.out = 1000)
r_values <- reg_fun(x_seq)   # 

lines(x_seq, r_values, type = 's', 
      col = adjustcolor("darkgreen", alpha.f = 0.4), 
      lwd = 3) 

# Add the regression function threshold
abline(h = 0.5, lty = 3, lwd =2)

# Add Legend
legend("topleft", legend = c("Y = 0", "Y = 1", "r(x)",'threshold'), 
       col = c('darkblue','darkred','darkgreen','black'), lty = c(0, 0, 1,3), 
       pch = c(16, 16, NA,NA), pt.cex = c(0.7, 0.7, NA,NA), lwd = c(NA, NA, 2,2))
```

This visualization shows the generated dataset and the regression
function $r(x)$ used to model the probability of $Y = 1$. The blue
points represent observations where the outcome is 0, and the red points
represent observations where the outcome is 1. The green stepwise line
represents the regression function $r(x)$, which models the posterior
probability $\mathbb{P}(Y = 1 | X = x)$ based on the underlying
distributions defined for $X$ given $Y$. The stepwise nature of the
function reflects how probabilities change sharply at the boundaries,
consistent with the uniform distributions specified in the data
generation process. The dotted black horizontal line at $r(x) = 0.5$
acts as a classification threshold. According to the Bayes classifier,
$\eta(x) = 1$ if and only if $r(x) > 0.5$. This decision rule splits the
feature space into two regions: predicted as $Y = 0$ where
$r(x) \leq 0.5$ and predicted as $Y = 1$ where $r(x) > 0.5$.

## 2.2

Now, we focus on the evaluation of the Bayes classifier.

```{r point 1.3, echo=TRUE}
# We evaluate its performance

# Predict using the Bayes classifier
data$y_pred <- bayes_classifier(data$x)     # function defined in 'Functions.R'

accuracy <- mean(data$y == data$y_pred) * 100
cat("Accuracy of Bayes Classifier: ", accuracy, "%\n",sep = '')

```

```{r, echo=TRUE}
cat("Size of Y = 0: ", sum(data$y == 0), "\n")
cat("Size of Y = 1: ", sum(data$y == 1), "\n")

```

Since the dataset is highly balanced, Accuracy is a reasonable metric
for evaluating the performance of the classifier.

An accuracy of 73.3% for the Bayes classifier means that **73.3% of all predictions match the true class labels** in the dataset. This value reflects the proportion of correctly classified instances out of the total number of observations.  

Since the Bayes classifier is theoretically optimal under the assumed probability distributions, this accuracy represents the best possible performance given the uniform distributions of \( X \mid Y \). However, this does not imply perfect classification. The classifier minimizes the overall error rate but does so by prioritizing specificity, which results in a high number of **false negatives** while avoiding **false positives** entirely.  

Thus, while 73.3% accuracy indicates strong performance, it must be interpreted alongside other metrics, such as recall and precision, to fully understand its classification behavior.


```{r, echo=TRUE}
# Compute Confusion Matrix
table_bayes <- table(Predicted = data$y_pred, Actual = data$y)

# Print Results
cat("Bayes Classifier Performance Metrics:\n")
cat("\nConfusion Matrix:\n")
print(table_bayes)

```

The misclassification pattern observed in the Bayes classifier stems
directly from the probabilistic properties of the data-generating
process. The conditional distributions of $X$ given $Y$ are defined as:

$$
(X \mid Y = 0) \sim \text{Unif}(-3,1), \quad (X \mid Y = 1) \sim \text{Unif}(-1,3)
$$

Both distributions are uniform over their respective intervals. Since
the Bayes classifier assigns $Y = 1$ when the posterior probability
$P(Y = 1 \mid X)$ exceeds $0.5$, the decision boundary is determined by:

$$
P(Y = 1 \mid X) = \frac{\pi_1 f_1(X)}{\pi_1 f_1(X) + \pi_0 f_0(X)}
$$

where $\pi_1 = P(Y = 1)$ and $\pi_0 = P(Y = 0)$ are the class priors,
both equal to $0.5$. Given that both conditional densities are uniform,
the decision threshold occurs at $X = 1$. This implies:

-   **For** $X > 1$, the classifier predicts $Y = 1$, as
    $P(Y = 1 \mid X) > 0.5$.\
-   **For** $X \leq 1$, the classifier predicts $Y = 0$, as
    $P(Y = 0 \mid X) \geq P(Y = 1 \mid X)$.

### Misclassification Analysis

The overlap between the two conditional distributions exists in the
range $[-1,1]$. Within this region, both classes have nonzero
probability, but the Bayes classifier **always predicts** $Y = 0$, as
the posterior probability favors class $Y = 0$. This leads to
**systematic misclassification of some** $Y = 1$ observations,
contributing to a high false negative rate.

### Confusion Matrix Breakdown

-   **False Negatives (267 cases):**\
    These occur because observations where $Y = 1$ and $X \in [-1,1]$
    are incorrectly classified as $Y = 0$.\
-   **True Positives (226 cases):**\
    These correspond to instances where $Y = 1$ and $X > 1$, which are
    correctly identified by the classifier. Given the uniform nature of
    $X \mid Y = 1$, approximately half of the $Y = 1$ cases fall within
    this range.\
-   **No False Positives:**\
    The classifier never predicts $Y = 1$ when the true label is
    $Y = 0$, ensuring perfect specificity but reducing recall.

```{r}
# Extract elements from the confusion matrix
TP <- table_bayes[2,2]
TN <- table_bayes[1,1]
FP <- table_bayes[2,1]
FN <- table_bayes[1,2]

# Compute Precision, Recall, and F1-score
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
f1_score <- 2 * (precision * recall) / (precision + recall)

# Print Results
cat("\nPrecision: ", round(precision, 4), "\n")
cat("Recall: ", round(recall, 4), "\n")
cat("F1-score: ", round(f1_score, 4), "\n")


```

#### 1. Precision = 1 (100%)

-   Precision is the proportion of correctly predicted positives
    ($Y = 1$) out of all instances predicted as $Y = 1$.\
-   A precision of 100% means that every time the classifier predicts
    $Y = 1$, it is always correct.\
-   This aligns with the confusion matrix, where there are zero false
    positives ($FP = 0$)—the classifier never predicts $Y = 1$ when the
    true class is $Y = 0$.\
-   High precision indicates that the classifier is conservative in
    assigning $Y = 1$ and only does so when it is very confident.

#### 2. Recall = 0.4584 (45.84%)

-   Recall measures how many of the actual $Y = 1$ cases were correctly
    identified by the classifier.\
-   A recall of 45.84% means that more than half of the true $Y = 1$
    cases were missed, being classified as $Y = 0$ instead.\
-   This is a direct consequence of the conservative nature of the Bayes
    classifier, which only predicts $Y = 1$ when $X > 1$.\
-   Since true $Y = 1$ values exist in the overlapping range
    $X \in [-1,1]$, they are misclassified as $Y = 0$, leading to a high
    number of false negatives ($FN = 267$).

#### 3. F1-score = 0.6287 (62.87%)

-   The F1-score is the harmonic mean of precision and recall, balancing
    both metrics.\
-   A score of 62.87% reflects a trade-off between the perfect precision
    (which avoids false positives) and the relatively low recall (which
    misses many true positives).\
-   The moderate F1-score suggests that while the classifier is highly
    conservative, it sacrifices recall significantly.

```{r}
# Load required library
library(pROC)

# Compute ROC and AUC
roc_curve <- roc(data$y, data$y_pred)
auc_value <- auc(roc_curve)

# Plot ROC Curve
plot(roc_curve, col = "blue", lwd = 2, main = "ROC Curve - Bayes Classifier",
     xlab = "False Positive Rate (1 - Specificity)", ylab = "True Positive Rate (Sensitivity)")

# Add reference line for random guessing
abline(a = 0, b = 1, lty = 2, col = "red")  

# Add legend
legend("bottomright", legend = c("Bayes Classifier", "Random Guessing"),
       col = c("blue", "red"), lty = c(1, 2), lwd = c(2, 2), bty = "n")

# Print AUC
cat("\nArea Under the Curve (AUC): ", round(auc_value, 4), "\n")




```

### ROC Curve Analysis for the Bayes Classifier

The ROC curve illustrates the trade-off between **sensitivity** (True
Positive Rate, TPR) and **false positive rate** (FPR) for the Bayes
classifier. The classifier's ability to distinguish between the two
classes is evaluated based on this curve, where an ideal classifier
would approach the upper-left corner of the plot.

------------------------------------------------------------------------

### Interpretation of the ROC Curve

-   The **blue line** represents the performance of the Bayes
    classifier.\
-   The **gray diagonal** represents a random classifier, serving as a
    baseline.\
-   The **red dashed line** corresponds to a completely random
    classifier with no predictive power.

The **x-axis** represents the **false positive rate (FPR)**, given by:

$$
FPR = \frac{FP}{FP + TN} = 1 - \text{Specificity}
$$

where $FP$ (false positives) are negative cases incorrectly predicted as
positive, and $TN$ (true negatives) are correctly predicted negative
cases.

The **y-axis** represents the **true positive rate (TPR)** or
**sensitivity**, defined as:

$$
TPR = \frac{TP}{TP + FN}
$$

where $TP$ (true positives) are correctly classified positive cases, and
$FN$ (false negatives) are actual positives misclassified as negatives.

The **initial segment of the curve remains at the bottom-left corner**,
indicating that the classifier is highly conservative—it rarely predicts
$Y = 1$, leading to a low sensitivity at the beginning. As the threshold
shifts, the sensitivity increases rapidly, reflecting that once the
classifier starts predicting $Y = 1$, it correctly identifies more true
positives. However, its initial reluctance results in a significant
number of **false negatives**.


------------------------------------------------------------------------

### Area Under the Curve (AUC)

The **Area Under the Curve (AUC)** quantifies the overall performance of
the classifier. AUC values range from **0 to 1**:

-   $AUC = 1$: Perfect classification.\
-   $AUC = 0.5$: Performance equivalent to random guessing.\
-   $AUC < 0.5$: Worse than random guessing (a classifier performing in
    the opposite direction).

Given the shape of the ROC curve, the **AUC is moderate**, suggesting
that while the classifier achieves **high specificity**, it suffers from
**low recall** due to its conservative nature.

The classifier excels in **avoiding false positives** but fails to
capture a significant number of **true positives**, making it **highly
precise but limited in recall**.

------------------------------------------------------------------------

## 2.3 Data-Driven Classifier

Here, we consider a comparison between the Bayes classifier and the
Logistic regression classifier.

Logistic regression is a supervised machine learning algorithm used for
binary classification problems. It models the probability that an
observation belongs to the positive class, $Y = 1$, given input
features, $X_1, \ldots, X_p$. The probability is modeled using the
logistic (sigmoid) function:

$$
\mathbb{P}(Y = 1 | X) = \frac{1}{1 + e^{-z}}
$$

where $z = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p$ is
a linear combination of input features, and $\beta$ are parameters
estimated using Maximum Likelihood Estimation (MLE). Predictions are
made by assigning $Y = 1$ if $\mathbb{P}(Y = 1 | X) > 0.5$ and $Y = 0$
otherwise. Logistic regression assumes a linear relationship between the
features and the log-odds, independent observations, and no
multicollinearity.

```{r}
set.seed(54)

# Train Set
train_data <- data

# Test Set
m = 500
test_data <- simulate(m)

# Logistic Regression Model
logistic_model <- glm(y ~ x, data = train_data, family = binomial)

# We predict probabilities on the test dataset
test_data$y_prob <- predict(logistic_model, newdata = test_data, type = "response")

#  And convert those probabilities to binary predictions using a 0.5 threshold
test_data$y_pred <- ifelse(test_data$y_prob > 0.5, 1, 0)

# Compute accuracy
test_accuracy <- mean(test_data$y == test_data$y_pred) * 100
cat("Test Accuracy of Logistic Regression: ", test_accuracy, "%\n")



```

Now, we evaluate the Bayes classifier on the test set:

```{r}
# Apply the Bayes classifier on the test set
test_data$bayes_pred <- bayes_classifier(test_data$x)  # Function must be defined in 'Functions.R'

# Compute Accuracy for Bayes Classifier
bayes_accuracy <- mean(test_data$y == test_data$bayes_pred) * 100
cat("Test Accuracy of Bayes Classifier:", bayes_accuracy, "%\n")

# Compare with Logistic Regression
cat("Test Accuracy of Logistic Regression:", accuracy, "%\n")

```

The Bayes classifier achieves slightly higher accuracy as it directly
leverages the true underlying data-generating process, serving as the
optimal benchmark for classification in this setup. In contrast,
logistic regression estimates the decision boundary from the data,
introducing variability and a minor performance drop due to practical
challenges in model estimation and potential misalignment with the true
data distribution. However, accuracy is not the only performance metric and we will further explore the comparison between the two classifiers.

```{r}
# We define a sequence of x values to generate predictions over a smooth range,
# ensuring a detailed representation of the regression function.
x_seq <- seq(min(data$x), max(data$x), length.out = 1000)

# We compute the predicted probabilities using our logistic regression model.
# This allows us to visualize how the model estimates the probability of Y = 1 across the x-axis.
logistic_prob <- predict(logistic_model, newdata = data.frame(x = x_seq), type = "response")

# We plot the dataset, differentiating between the two classes.
# Points corresponding to Y = 1 are shown in a semi-transparent dark red,
# while points for Y = 0 are displayed in a semi-transparent dark blue.
plot(data$x, data$y, 
     col = ifelse(data$y == 1, adjustcolor("darkred", alpha.f = 0.1), 
                                  adjustcolor("darkblue", alpha.f = 0.1)), 
     pch = 16, cex = 0.7, xlab = "X", ylab = "Y", 
     main = "Generated Dataset and Regression Function", 
     sub = expression(eta(x) == 1 ~ "if and only if" ~ r(x) > 0.5))

# We compute the Bayes regression function, which represents the theoretical 
# optimal probability of Y = 1 given X. This function should be pre-defined in 'Functions.R'.
r_values <- reg_fun(x_seq)

# We overlay the Bayes regression function onto the plot using a step function.
# This serves as the optimal decision boundary for classification.
lines(x_seq, r_values, type = 's', col = adjustcolor("darkgreen", alpha.f = 0.6), lwd = 3) 

# We add the estimated probabilities from the logistic regression model.
# This allows us to compare our logistic model's estimated function against the Bayes optimal function.
lines(x_seq, logistic_prob, col = "purple", lwd = 3, lty = 2)

# We add a legend to clarify the different elements present in the plot.
# - Y = 0 and Y = 1 represent the data points.
# - The Bayes Regression Function represents the optimal probability function.
# - The Logistic Regression Estimate shows the model's learned probabilities.
legend('bottomright', 
       legend = c("Y = 0", "Y = 1", "Bayes Regression Function", "Logistic Regression Function"),
       col = c("darkblue", "darkred", "darkgreen", "purple"), 
       lty = c(NA, NA, 1, 2), 
       pch = c(16, 16, NA, NA), 
       pt.cex = c(1.2, 1.2, NA, NA),  # We increase the point size for better visibility.
       lwd = c(NA, NA, 3, 3),          # We ensure the regression function lines are clearly visible.
       bty = "n")  # We remove the legend box for a cleaner aesthetic.



```

The Bayes function represents the **optimal classifier**, showing the
true probability $P(Y = 1 \mid X)$ with a **stepwise shape**, reflecting
an abrupt decision boundary. In contrast, logistic regression estimates
probabilities using a **smooth, sigmoidal function**, making it a
practical but imperfect approximation. The dataset shows a **clear
separation** between classes, making classification straightforward.
While the Bayes function is **ideal but unknown**, logistic regression
is **robust** and generalizes well, though it may struggle with sharp
decision boundaries. This highlights the trade-off between **theoretical
optimality and practical modeling**, as logistic regression smooths
transitions that might be abrupt in reality.

## Performance Analysis of the Logistic Regression Classifier  

Logistic regression provides a **data-driven approach** to classification by estimating the decision boundary directly from the training data. Unlike the Bayes classifier, which uses **true conditional distributions**, logistic regression models the probability \( P(Y = 1 \mid X) \) using a **sigmoid function**, making it a more flexible but approximate solution.

---

### 1. Evaluation Metrics  

To compare the classifiers beyond accuracy, we compute the **precision, recall, and F1-score** for logistic regression.

```{r}
# Compute Confusion Matrix for Logistic Regression
table_logistic <- table(Predicted = test_data$y_pred, Actual = test_data$y)

# Extract elements from confusion matrix
TP_log <- table_logistic[2,2]
TN_log <- table_logistic[1,1]
FP_log <- table_logistic[2,1]
FN_log <- table_logistic[1,2]

# Compute Precision, Recall, and F1-score
precision_log <- TP_log / (TP_log + FP_log)
recall_log <- TP_log / (TP_log + FN_log)
f1_score_log <- 2 * (precision_log * recall_log) / (precision_log + recall_log)

# Print results
cat("\nLogistic Regression Performance Metrics:\n")
cat("Precision: ", round(precision_log, 4), "\n")
cat("Recall: ", round(recall_log, 4), "\n")
cat("F1-score: ", round(f1_score_log, 4), "\n")

# Print Confusion Matrix
cat("\nConfusion Matrix for Logistic Regression:\n")
print(table_logistic)

# Extract elements from the confusion matrix
TP <- table_bayes[2,2]
TN <- table_bayes[1,1]
FP <- table_bayes[2,1]
FN <- table_bayes[1,2]


cat("\nBayes Classifier Performance Metrics:\n")
# Compute Precision, Recall, and F1-score
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
f1_score <- 2 * (precision * recall) / (precision + recall)

# Print Results
cat("\nPrecision: ", round(precision, 4), "\n")
cat("Recall: ", round(recall, 4), "\n")
cat("F1-score: ", round(f1_score, 4), "\n")

table_bayes <- table(Predicted = data$y_pred, Actual = data$y)

# Print Results
cat("Bayes Classifier Confusion Matrix:\n")
cat("\nConfusion Matrix:\n")
print(table_bayes)

```
```{r}
# Compute ROC and AUC for Logistic Regression
roc_curve_logistic <- roc(test_data$y, test_data$y_prob)
auc_logistic <- auc(roc_curve_logistic)
# Compute ROC and AUC for Bayes Classifier
roc_curve_bayes <- roc(test_data$y, test_data$bayes_pred)
auc_bayes <- auc(roc_curve_bayes)

# Plot ROC Curve
plot(roc_curve_logistic, col = "purple", lwd = 2, main = "ROC Curves: Logistic Regression vs. Bayes",
     xlab = "False Positive Rate (1 - Specificity)", ylab = "True Positive Rate (Sensitivity)")
lines(roc_curve_bayes, col = "blue", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "red")  # Reference Line

# Add legend
legend("bottomright", legend = c("Logistic Regression", "Bayes Classifier", "Random Guessing"),
       col = c("purple", "blue", "red"), lty = c(1, 1, 2), lwd = c(2, 2, 2), bty = "n")

# Print AUC values
cat("\nArea Under the Curve (AUC) - Logistic Regression: ", round(auc_logistic, 4), "\n")
cat("Area Under the Curve (AUC) - Bayes Classifier: ", round(auc_bayes, 4), "\n")

```
### Performance Comparison  

Logistic regression achieves a precision of 0.7719, meaning that approximately 77% of the observations predicted as \( Y = 1 \) are correctly classified. In contrast, the Bayes classifier has a perfect precision of 1.0, indicating that it never incorrectly predicts \( Y = 1 \). This occurs because the Bayes classifier is highly conservative, only predicting \( Y = 1 \) when it is certain, completely avoiding false positives.  

Recall is significantly higher for logistic regression at 0.7184 compared to 0.4584 for the Bayes classifier. This suggests that logistic regression captures a larger proportion of true \( Y = 1 \) cases, whereas the Bayes classifier frequently misclassifies them as \( Y = 0 \), leading to a high number of false negatives.  

The F1-score, which balances precision and recall, is 0.7442 for logistic regression, whereas the Bayes classifier achieves 0.6287. The lower F1-score of the Bayes classifier is a direct result of its low recall, despite having perfect precision. Logistic regression, by maintaining a better balance between precision and recall, achieves a more effective overall performance.  

Examining the confusion matrices, the Bayes classifier results in **507 true negatives, 267 false negatives, 0 false positives, and 226 true positives**. The complete absence of false positives explains its perfect precision. However, the 267 false negatives highlight its conservative nature, as it avoids predicting \( Y = 1 \) unless highly certain. Logistic regression produces **203 true negatives, 69 false negatives, 52 false positives, and 176 true positives**. While it has more false positives than the Bayes classifier, it compensates with significantly fewer false negatives, improving its recall.  

The test accuracy of the Bayes classifier is **75.4%**, slightly higher than the **73.3%** accuracy of logistic regression. While this suggests that the Bayes classifier minimizes overall misclassification slightly better, accuracy alone does not capture the trade-offs between false positives and false negatives.  

The ROC curve further demonstrates the differences between the classifiers. The curve for logistic regression consistently stays above that of the Bayes classifier, indicating better sensitivity across different classification thresholds. The area under the curve (AUC) for logistic regression is **0.8758**, significantly higher than the Bayes classifier’s **0.749**, confirming that logistic regression has a stronger ability to differentiate between the two classes.  

Overall, the Bayes classifier adopts a conservative strategy, ensuring no false positives but suffering from a high number of false negatives. Logistic regression, being more flexible, finds a better balance between false positives and false negatives. Given its superior AUC and recall, logistic regression provides a more effective classification performance in this context, despite the Bayes classifier achieving a marginally higher accuracy.


## 3. Repeated Sampling

In this section, we conduct a repeated sampling experiment to compare
the performance of the Bayes classifier and logistic regression.
Specifically, we simulate $M = 10,000$ datasets, each consisting of a
training set with $n = 1,000$ observations and a test set with $m = 500$
observations, drawn from the specified data-generating process.

For each simulated dataset:

1.  **The Bayes classifier** is applied to the test set, and its
    accuracy is recorded. This serves as the theoretical benchmark since
    it directly utilizes the known underlying distributions.

2.  **Logistic regression** is trained on the training set and evaluated
    on the test set. The accuracy of its predictions on the test set is
    recorded, representing the practical performance of a data-driven
    classifier.

After $M$ repetitions, we compute the mean and standard deviation of the
accuracies for both classifiers to assess their overall performance.

```{r, echo=TRUE}
set.seed(78)  # Ensure reproducibility
M <- 10000  # Number of iterations
n <- 1000   # Train set size
m <- 500    # Test set size
bayes_accuracies <- numeric(M)
logistic_accuracies <- numeric(M)

for (i in 1:M) {
  # Generate training dataset
  train_data <- simulate(n)
  
  # Train Logistic Regression
  logistic_model <- glm(y ~ x, data = train_data, family = binomial)
  
  # Generate test dataset (new independent sample)
  test_data <- simulate(m)
  
  # Bayes Classifier Predictions on test data
  test_data$y_pred_bayes <- bayes_classifier(test_data$x)
  bayes_accuracies[i] <- mean(test_data$y == test_data$y_pred_bayes)
  
  # Logistic Regression Predictions on test data
  prob_pred <- predict(logistic_model, newdata = test_data, type = "response")
  test_data$y_pred_logistic <- ifelse(prob_pred > 0.5, 1, 0)
  logistic_accuracies[i] <- mean(test_data$y == test_data$y_pred_logistic)
}

# Compute average accuracies
bayes_mean_acc <- mean(bayes_accuracies) * 100
logistic_mean_acc <- mean(logistic_accuracies) * 100

cat("Average Accuracy over", M, "iterations:\n")
cat("Bayes Classifier:", bayes_mean_acc, "%\n")
cat("Logistic Regression:", logistic_mean_acc, "%\n")





```

The results of the repeated sampling experiment indicate that logistic
regression marginally outperforms the Bayes classifier. The **Bayes
classifier** achieves a mean accuracy of approximately 75.03%, while
**logistic regression** attains a slightly higher mean accuracy of
75.04%.

At first glance, this slight advantage for logistic regression may seem
unexpected, given that the Bayes classifier is theoretically optimal
under the assumption of a perfectly known data-generating process.
However, this result highlights an important practical consideration:
while the Bayes classifier relies entirely on the assumed distributions
of $X \mid Y$, logistic regression learns the decision boundary directly
from the data. This flexibility allows logistic regression to adapt to
small variations or noise in the simulated datasets that may not
perfectly conform to the theoretical assumptions.

Additionally, the data-driven nature of logistic regression enables it
to generalize effectively, even in cases where the underlying
distributions are subject to slight deviations. This adaptability gives
logistic regression a practical advantage in real-world applications,
where the true data distribution is often unknown or imperfectly
specified.

Overall, these findings reinforce the robustness of logistic regression
as a versatile classifier, capable of performing on par with or even
slightly surpassing theoretical benchmarks like the Bayes classifier in
applied settings.

# Part 2

In this section, we aim to summarize and present the core methodology
described by Jerome H. Friedman in his paper on multivariate
goodness-of-fit and two-sample testing. Our approach is to distill the
key steps into a clear and well-structured pseudo-code, providing a
high-level overview of the testing procedure.

We begin by outlining how to generate and process the datasets, followed
by leveraging a binary classifier to assign scores that differentiate
the two samples. These scores are then analyzed using univariate
two-sample tests, such as Mann-Whitney or Kolmogorov-Smirnov, which act
as baseline methods for assessing distributional differences. Finally,
we describe how the null hypothesis is tested and interpreted to draw
meaningful conclusions.

Our goal is to highlight how machine learning techniques, particularly
binary classifiers, can be effectively combined with traditional
statistical methods to address complex multivariate testing problems.
This pseudo-code is intended to serve as a guide for implementing the
procedure in practice while ensuring clarity and precision.

## Pseudo-code for Friedman's paper

------------------------------------------------------------------------

1.  Generate Datasets:
    -   Start with two datasets:
        -   Sample 1: $\{x_i\}$ with $N$ observations drawn from $p(x)$.
        -   Sample 2: $\{z_i\}$ with $M$ observations drawn from $q(x)$.
2.  Combine the Datasets:
    -   Merge the two datasets into one:
        -   $\{u_i\} = \{x_i\} \cup \{z_i\}$.
    -   Assign labels to identify the origin of each observation:
        -   Assign $y_i = 1$ for data points from Sample 1 ($\{x_i\}$).
        -   Assign $y_i = -1$ for data points from Sample 2 ($\{z_i\}$).
3.  Train a Binary Classifier:
    -   Use a machine learning model (e.g. logistic regression) to
        classify the combined data.
    -   Input: The dataset $\{u_i\}$.
    -   Output: Scores $\{s_i\}$ for each data point, where
        $s_i = F(u_i)$ is the score assigned by the classifier.
4.  Compute Scores for Each Sample:
    -   Split the scores into two groups based on their original
        datasets:
        -   $S^+ = \{s_i \, \text{for all } x_i \, \text{(Sample 1)}\}$.
        -   $S^- = \{s_i \, \text{for all } z_i \, \text{(Sample 2)}\}$.
5.  Perform a Two-Sample Test:
    -   Apply a univariate two-sample test (like Mann-Whitney or
        Kolmogorov-Smirnov) on the scores:
        -   Compute the test statistic: $t = T(S^+, S^-)$.
6.  Test the Null Hypothesis:
    -   Estimate the null distribution for the test statistic:
        -   If different datasets are used for training and scoring:
            -   Use the standard null distribution (e.g., for
                Mann-Whitney or Kolmogorov-Smirnov).
        -   If the same dataset is used for both training and scoring:
            -   Use a permutation test to create the null distribution.
    -   Compare the observed test statistic $t$ to the critical value
        for a chosen significance level $\alpha$.
    -   Reject the null hypothesis ($p = q$) if $t$ is greater than the
        critical value.
7.  Draw a Conclusion:
    -   Based on the significance test, decide whether to reject or fail
        to reject the null hypothesis.
    -   If rejected, conclude that the two samples are likely drawn from
        different distributions.

------------------------------------------------------------------------

Friedman suggests using the **Mann-Whitney** and **Kolmogorov-Smirnov**
(KS) tests as baseline methods because they are simple, reliable, and
effective for comparing two distributions. Both tests are **univariate**
and **non-parametric**, which means they don’t rely on the data
following a specific distribution. This makes them flexible and
applicable in many different situations.

The Mann-Whitney test is particularly helpful when we want to compare
the central tendencies, such as the medians, of two groups. It works by
ranking all the observations and comparing the ranks between the two
samples, making it ideal for identifying whether the central values of
the groups differ. In contrast, the KS test looks at the overall shape
of the distributions by comparing their cumulative distribution
functions (CDFs). This allows it to detect differences in not just
central tendencies but also in spread, shape, or location, providing a
broader perspective.

These tests are great starting points because they are **easy to
understand and implement**. They don’t require complicated calculations,
and their **non-parametric nature** makes them adaptable to different
types of data **without needing strict assumptions**. Since Friedman's
method generates univariate scores from the classifier, these tests are
perfectly suited to evaluate those scores, ensuring they align well with
the problem at hand.


